An open-source \old{1D}\new{full 3D} electromagnetic modeler \new{for 1D VTI media} in Python: empymod

### 2017-05-30 ManuscriptGeophysics2016.tex
Draft of new revision

% \documentclass[manuscript]{geophysics}
% \documentclass[paper,twocolumn,twoside]{geophysics}
% \documentclass[paper]{geophysics}
\documentclass[manuscript,revised]{geophysics}
%%fakesection ===    PACKAGES & DEFINITIONS    ===

@@

\title{An open-source 1D electromagnetic modeler in Python: empymod}

@@

\ms{GEO-2016-0626}

@@

\footer{}
\lefthead{Werthmüller}
\righthead{Open-source 1D EM modeler in Python}

@@

%%fakesection ===    ABSTRACT    ===
\begin{abstract} % 1-2 sentence(s) each
  The presented Python-code empymod computes the three-dimensional
  electromagnetic field in a layered-earth with vertical transverse isotropy by
  combining and extending two earlier presented algorithms in this journal.
  The bottleneck in frequency- and time-domain calculations of electromagnetic
  responses derived in the wavenumber-frequency domain is the transformations
  from wavenumber to space domain and from frequency to time domain, the
  so-called Hankel and Fourier transforms. Three different Hankel transform
  methods (quadrature, quadrature-with-extrapolation, and filters) and four
  different Fourier transform methods (FFT, FFTLog,
  quadrature-with-extrapolation, and filters) are included in empymod, which
  allows to compare these different methods in terms of speed and precision.
  The best transform in terms of speed and precision depends on the modeled
  frequencies: available filters, for instance, are very fast and precise for
  frequencies in the range of controlled-source electromagnetic data, but fail
  in the frequency range of ground penetrating radar. Conventional quadrature,
  on the other hand, is in comparison very slow but can model any frequency.
  Examples comparing empymod with analytical solutions and with existing
  electromagnetic modelers illustrate the capabilities of empymod.
\end{abstract}

@@

To my knowledge, empymod is the first code that is freely available which
calculates the full wavefield for a layered-earth model with vertical isotropy
and has different types of Hankel transforms (QWE, QUAD, FHT) and various
methods for the Fourier transform (FFTLog, sine/cosine-filters, QWE, FFT). An
interesting use-case for empymod is therefore comparison studies of the
different methods, up to building hybrid inversion schemes. The code is
published under the lax permissive \emph{Apache Version 2.0 license}, which
makes it available to everyone for free, even for commercial purposes. It
therefore might prove to be valuable for students and professionals alike. It
is written in Python, a modern, cross-platform, free and open-source
programming language. With its scientific libraries, mainly the numeric and
scientific modules \texttt{NumPy} and \texttt{SciPy}, it creates an extremely
powerful numerical calculation stack: Even though it is an interpreted language
it can be very fast, as \texttt{NumPy}/\texttt{SciPy} use under the hood
routines in Fortran or in C (using for instance the BLAS/Lapack libraries) for
the linear algebra computations, and Python is merely the glue. The comparisons
show that empymod is as fast as the existing codes. The code is vectorized
(avoiding loops) as much as possible, which improves computation. The most
time-consuming calculations have furthermore a flag to run parallelized.
However, on a larger scale such as an inversion routine one would probably want
to parallelize the kernel calls instead of the kernel itself. The code is
hosted on GitHub, which makes it easy for anyone to improve and contribute to
the code: \url{https://github.com/prisae/empymod}. The documentation is hosted
on \url{https://empymod.readthedocs.io}. \cite{TLE.17.Werthmuller} gives a
tutorial-style introduction to EM modeling using empymod.

@@

The code consists of 5 files; these are 3 core modules plus \emph{utils}, which
contains input checks and other utilities, and \emph{filters}, containing the
FHT filter coefficients. The three core routines are: (1) \emph{kernel}, where
the wavenumber-domain calculation is carried out; (2) \emph{transform}, where
the Hankel and Fourier transforms are computed; and (3) \emph{model}, which
contains the actual modeling routines for end users. The main modeling routine
is \emph{bipole}, which can calculate frequency- and time-domain responses for
arbitrary oriented, electric or magnetic bipole sources and receivers of finite
length.

The wavenumber-domain calculation in \emph{kernel} follows
\cite{GEO.15.Hunziker}, and calculates as such the complete wavefield for a
layered VTI model. The code makes no assumptions about the model: Source and
receiver can be placed anywhere in the model, including first and last layer,
and you have to define if the first layer is air or not. Bipoles can cross
layer boundaries. Depths, frequencies, and source-receiver configuration have
to be defined, and each layer is characterized with its horizontal resistivity
$\rho_\textrm{h}$, its electrical anisotropy $\lambda$, where $\lambda =
\sqrt{\rho_\textrm{v}/\rho_\textrm{h}}$, its horizontal and vertical magnetic
permeabilities $\mu_\mathrm{h}$ and $\mu_\mathrm{v}$, and the horizontal and
vertical electric permittivities $\epsilon_\mathrm{h}$ and
$\epsilon_\mathrm{v}$. The main difference between EMmod and empymod is the
Hankel transform, and a few additional fundamental differences: EMmod uses 2nd
order Bessel functions of the first kind $J_2$. Published FHT filters
generally provide coefficients for 0th and 1st order Bessel functions $J_0,
J_1$ only. Therefore, the recurrence relation
%
\begin{equation}
  J_2(kr) = \frac{2}{kr}J_1(kr) - J_0(kr)
  \label{eq:j2}
\end{equation}
%
is used, where $k$ and $r$ are the wavenumber and space-domain parameters,
respectively. Other differences are that empymod includes Fourier transforms,
hence time-domain calculation, and can model arbitrary rotated, finite bipoles.

@@

Calculation time can be reduced by using the splined version of the QWE and the
splined and lagged versions of the FHT. In addition, all time-consuming
calculations are set up to be able to run in parallel, with the help of the
Python-module numexpr. This can significantly speed up calculations if you run
big models with many layers and for many offsets and frequencies. However, if
you include empymod in an inversion scheme then it might be better to
parallelize the calls to empymod, instead of empymod itself.

It is important to note that calculations in the wavenumber domain depend
\emph{only} on offset $r = \sqrt{x^2+y^2}$; the scaling factor, which depends
on the angle $\varphi = \rm{atan2}(y, x)$, is multiplied afterwards. In order
to calculate for instance a circle around the source, the kernel has to be
called only once, and subsequently scaled by the angle-dependent factor for
each source-receiver pair. This makes the splined version very powerful, as it
can be used for irregularly distributed data as long as all receiver depths are
the same.

@@

@@

Table~\ref{tbl:analytical} shows run times for these models. A few notes worth
mentioning: empymod carries out a lot of input checks, as it is quite forgiving
in what you input. EMmod, on the other hand, reads the input file and writes
the result to an output file, and copies the first quadrant to the other three.
Both have therefore some overhead, and the comparison is not strictly 1:1. For
the comparison the same model was used as shown above, on a regular grid with a
spacing of 100\,m. On the test-machine the standard QWE and FHT empymod would
run into memory issues for denser spacing, hence arrays of more than some
10\,000s of offsets. The splined and lagged versions of EMmod could handle it,
however.
%
\tabl[btp]{analytical}{Run times for the half-space model shown in Figure~1, on
  a regular grid with spacing of 100\,m, for EMmod and different Hankel
  transform settings of empymod (times in milliseconds).}{
  \centering
  \begin{tabularx}{320pt}{rAAABBB}
  \toprule%
  EMmod & \multicolumn{3}{c}{QWE} & \multicolumn{3}{c}{FHT} \\
        & 1 & 2 & 3               & 1 & 2 & 3 \\
  \midrule%
   5040 & 10300 & 2530 & 1210 & 1480 & 875 & 6 \\
  \bottomrule%
  \end{tabularx}
}
%
QWE becomes faster for decreasing precision, not surprisingly, and the splined
and lagged version of FHT are faster than the standard version. The lagged FHT
is the fastest by quite a margin, and about 1/3 of that time is from the input
checks, so it takes only a few milliseconds to calculate the 11\,025 offsets.
What is interesting here is that the lagged FHT is still more precise than
EMmod with the given settings, which means a speed-up of a factor 1000 for the
same result.

Many 1D CSEM codes use the diffusive approximation that is valid for low
frequencies. The appealing part of the derivation of \cite{GEO.15.Hunziker} is
that it models the complete wavefield, hence it is also valid for high
frequencies and therefore wave-phenomena. As an extreme case,
\cite{GEO.15.Hunziker} show a 1D example for ground-penetrating radar with a
center frequency of 250\,MHz. To do so, they calculate 2048 frequencies in the
range of 1\,MHz to 2048\,MHz for the Fast Fourier transform from frequency to
time domain. Here I only calculated the FFT with 850 frequencies from 1\,MHz
to 850\,MHz and then zero-padded up to 2048\,MHz, for EMmod and empymod; tests
have shown that this is sufficient. Figure~\ref{fig:gpr} shows the model with
the survey setup and the four results for EMmod, QUAD for relative and absolute
tolerance of $10^{-12}$ and $10^{-20}$, respectively, empymod with a 21\,pt QWE
and relative and absolute tolerance of $10^{-10}$ and $10^{-18}$, respectively,
and empymod with FHT with the filter Key09-401, together with the analytical
solution for the arrival of the direct wave (red), the wave refracted at the
surface (yellow), and the wave reflected at the subsurface interface (green).

EMmod and the QUAD and QWE transforms of empymod yield pretty much the same
result. It is only at later times that some noise can be seen, different in
each method. FHT shows the first arrivals well, however, afterwards the result
becomes extremely noisy.
%
\plot*[btp]{gpr}{width=1\textwidth}{GPR example for (a) EMmod, (b) QUAD
  for relative and absolute tolerance of $10^{-12}$ and $10^{-20}$,
  respectively, (c) empymod with 51\,pt QWE for relative and absolute tolerance
  of $10^{-8}$ and $10^{-15}$, respectively, and (d) empymod with FHT
  (Key09-401); run times are given in the subplot titles.}
%
Interesting are the calculation times for these three results. EMmod took
roughly 9.5 hours to calculate, empymod with QUAD about 7.8 hours, empymod with
QWE about 4.4 hours, and empymod with FHT under a minute. Note that QWE has
internally a check: If the kernel is too steep within given intervals, it uses
QUAD instead, as QWE cannot handle very steep functions. QWE uses in about 1/3
% (37%)
of the calculation of this example QUAD. Surprising is how good the FHT result
is, given that the filter was designed for CSEM data, hence frequencies 6 to 9
orders of magnitudes smaller and much bigger offsets. In any case,
EMmod/empymod are not optimized for nor intended to model GPR data, and
calculations in time-domain will be much faster and more precise. It is
nevertheless an interesting proof of concept and shows that EMmod/empymod model
the entire EM field. This leaves the question if it would be possible to
create a filter that works for the frequencies and the offsets required for GPR
calculations.

@@

\cite{GEO.12.Key} not only introduces QWE to geophysics, but also compares QWE
to FHT for various filters, some of which were specifically created for the
comparison. In order to calculate the models he translates parts of Dipole1D
(\cite{GEO.09.Key}) from Fortran to Matlab. The models are therefore isotropic
and use the diffusive approximation. He summarizes succinct and clear the FHT
method, outlines the QWE method, and made the codes available from the SEG
website.

@@

Table~\ref{tbl:runtimes} lists the run times for exactly the same models as
\cite{GEO.12.Key} compared in his Table~1: 1\,km water layer, followed by a
1\,km overburden of 1\,$\Omega\,$m, 100\,m reservoir of 100\,$\Omega\,$m, and
lastly 1 or 96 layers of underburden with a resistivity of 1\,$\Omega\,$m;
frequency is 1\,Hz, source depth 990\,m, and receivers are on the sea-bottom;
there are 1, 5, 21, 81, or 321 offsets between 0.5 and 20\,km. In this
comparison, a 9\,pt QWE is compared to a 201\,pt (Key12-201) and a 801\,pt
filter (Anderson82-801), where the absolute and relative tolerance are set so
that the QWE achieves a similar error-level as the filters ($\varepsilon_r =
10^{-6}$ for the standard QWE and $10^{-2}$ for the splined version,
$\varepsilon_a = 10^{-24}$ in both cases). In order to make a fair comparison I
re-run the script from \cite{GEO.12.Key}, and the run times on the test-machine
are roughly twice as fast as in the original paper. Next to it are the results
from empymod. It can be seen from the results that empymod is significantly
faster.%
%
\tabl[btp]{runtimes}%
{Run times comparing the codes from \cite{GEO.12.Key} (Key) and empymod (Wer)
for the same cases as in (Key), Table~1 (in milliseconds).}{
  \centering
\begin{tabularx}{1.05\textwidth}{ccYYWWYYWWYYWW}
  \toprule
  %
  & & \multicolumn{4}{c}{QWE: 9\,pt}&\multicolumn{4}{c}{FHT: 201\,pt filter}&
  \multicolumn{4}{c}{FHT: 801\,pt filter} \\
  %
  \cmidrule(rl){3-6} \cmidrule(rl){7-10} \cmidrule(rl){11-14}
  %
  & &
  \multicolumn{2}{>{\columncolor{MyGray}\arraybackslash}c}{Normal} &
  \multicolumn{2}{c}{Splined} &
  \multicolumn{2}{>{\columncolor{MyGray}\arraybackslash}c}{Normal} &
  \multicolumn{2}{c}{Lagged} &
  \multicolumn{2}{>{\columncolor{MyGray}\arraybackslash}c}{Normal} &
  \multicolumn{2}{c}{Lagged} \\
  %
  Lay& Off& Wer& Key& Wer& Key& Wer&  Key& Wer& Key& Wer&  Key& Wer& Key\\
  \midrule
  %
    5&   1&   7&  16&   2&  28&   1&   19&   1&  20&   2&   76&   2&  76\\
    5&   5&  13&  69&   4&  29&   3&   92&   1&  24&   7&  365&   3&  82\\
    5&  21&  19& 300&   6&  36&   8&  384&   2&  25&  25& 1522&   3&  83\\
    5&  81&  37&1154&  16&  62&  30& 1475&   2&  28& 100& 5865&   3&  86\\
    5& 321& 108&4581&  58& 165& 124& 5867&   2&  41& 437&23165&   3&  99\\
  \midrule
  100&  21& 118& 463&  13&  51&  92&  604&   8&  37& 304& 2402&  19& 127\\
  100&  81& 296&1792&  23&  75& 337& 2313&   8&  40&1234& 9215&  19& 129\\
  100& 321&1021&7226&  65& 178&1408& 9350&  10&  53&5448&36520&  19& 142\\
  %
  \bottomrule
\end{tabularx}}%
%
Important to note is, however, not the absolute speed of empymod, but the
difference between the various Hankel transform methods. The standard QWE
method, for a similar error level as FHT, is faster than the standard FHT mode
only for many offsets. More importantly, the lagged FHT is generally much
faster than any QWE. QWE is very useful to check the error level of a result.
If many kernel evaluations have to be carried out, and speed is of importance,
the lagged FHT is the preferred method. As \cite{GEO.12.Key} loops over each
wavenumber, the difference between the vectorized and non-vectorized version
can be best seen in the long 801\,pt filter.

@@

Table~\ref{tbl:runtimes2} lists the run times for the same model as before, but
for empymod and Dipole1D using the FHT method with the filter Key09-201, the
default filter in both Dipole1D and empymod. For the regular versions empymod
and Dipole1D perform very similar; empymod is a little faster for the smaller
tests, Dipole1D is slightly faster for the bigger tests. It can be seen from
the results that the parallel optimisation in empymod can result in a
significant speed-up. If the lagged convolution optimisation is chosen,
empymod is significantly faster than Dipole1D. In empymod, additional offsets
come at basically no cost in the lagged version, whereas Dipole1D still uses
more time to calculate more offsets. This is most likely due to the writing of
the output file, that becomes bigger with bigger offsets.
%
\tabl[btp]{runtimes2}
{Run times comparing Dipole1D (\cite{GEO.09.Key}) and empymod, using the filter
Key09-201 (default in Dipole1D; times in milliseconds). Optimisation: [-] None,
[par] parallel, [spl] lagged FHT.}{ \centering
  \begin{tabularx}{.5\textwidth}{ccAAABB}
  \toprule
  %
     &    & \multicolumn{3}{c}{empymod} & \multicolumn{2}{c}{Dipole1D}\\
  %
  \cmidrule(rl){3-5} \cmidrule(rl){6-7}
  %
  Lay& Off&  - & par & spl &   - & spl\\
  \midrule
  %
    5&   1&    1&   2& 1&    4&  4\\
    5&   5&    3&   3& 2&    6&  5\\
    5&  21&    9&   5& 2&   12&  8\\
    5&  81&   30&  10& 2&   36& 18\\
    5& 321&  124&  34& 2&  130& 56\\
  \midrule
  100&  21&   91&  53& 9&   90& 13\\
  100&  81&  340& 106& 9&  333& 23\\
  100& 321& 1423& 313& 9& 1321& 62\\
  %
  \bottomrule
\end{tabularx}}%
%
Dipole1D was compiled using the free and open-source \emph{gfortran} compiler.
Compiling it with the (proprietary) \emph{ifort} compiler might result in
slightly faster run times. Note that Dipole1D always calculates all six
receiver components, whereas empymod only calculates the required component(s).
On the other hand empymod computes the whole, anisotropic wavefield, whereas
Dipole1D is isotropic and uses the diffusive approximation.

@@

>>>>

% \documentclass[manuscript]{geophysics}
% \documentclass[paper,twocolumn,twoside]{geophysics}
\documentclass[paper]{geophysics}
% \documentclass[manuscript,revised]{geophysics}
%%fakesection ===    PACKAGES & DEFINITIONS    ===

@@

\title{An open-source \old{1D}\new{full 3D} electromagnetic modeler \new{for 1D VTI media} in Python: empymod}

@@

\ms{GEO-2016-0626.R2}

@@

\footer{}
\lefthead{Werthmüller}
\righthead{Open-source\old{ 1D} EM modeler in Python}

@@

%%fakesection ===    ABSTRACT    ===
\begin{abstract} % 1-2 sentence(s) each
  The presented Python-code empymod computes the three-dimensional
  electromagnetic field in a layered-earth with vertical transverse isotropy by
  combining and extending two earlier presented algorithms in this journal.
  The bottleneck in frequency- and time-domain calculations of electromagnetic
  responses derived in the wavenumber-frequency domain is the transformations
  from wavenumber to space domain and from frequency to time domain, the
  so-called Hankel and Fourier transforms. Three different Hankel transform
  methods (quadrature, quadrature-with-extrapolation, and filters) and four
  different Fourier transform methods (FFT, FFTLog,
  quadrature-with-extrapolation, and filters) are included in empymod, which
  allows to compare these different methods in terms of speed and precision.
  The best transform in terms of speed and precision depends on the modeled
  frequencies: \old{available filters}\new{published digital filters for the
  Hankel transform}, for instance, are very fast and precise for frequencies in
  the range of controlled-source electromagnetic data, but fail in the
  frequency range of ground penetrating radar. Conventional quadrature, on the
  other hand, is in comparison very slow but can model any frequency.  Examples
  comparing empymod with analytical solutions and with existing electromagnetic
  modelers illustrate the capabilities of empymod.
\end{abstract}

@@

To my knowledge, empymod is the first code that is freely available which
calculates the full wavefield for a layered-earth model with vertical isotropy
and has different types of Hankel transforms (QWE, QUAD, FHT) and various
methods for the Fourier transform (FFTLog, sine/cosine-filters, QWE, FFT). An
interesting use-case for empymod is therefore comparison studies of the
different methods, up to building hybrid inversion schemes. The code is
published under the lax permissive \emph{Apache Version 2.0 license}, which
makes it available to everyone for free, even for commercial purposes. It
\old{therefore} might \new{hence} prove to be valuable for students and
professionals alike, \new{and it is valuable for the objectives of reproducible
research.} It is written in Python, a modern, cross-platform, free and
open-source programming language. With its scientific libraries, mainly the
numeric and scientific modules \texttt{NumPy} and \texttt{SciPy}, it creates an
extremely powerful numerical calculation stack: Even though it is an
interpreted language it can be very fast, as \texttt{NumPy}/\texttt{SciPy} use
under the hood routines in Fortran or in C (using for instance the BLAS/Lapack
libraries) for the linear algebra computations, and Python is merely the glue.
The comparisons show that empymod is as fast as the existing codes. The code is
vectorized (avoiding loops) as much as possible, which improves computation.
The most time-consuming calculations have furthermore a flag to run
parallelized.  However, on a larger scale such as an inversion routine one
would probably want to parallelize the kernel calls instead of the kernel
itself. The code is hosted on GitHub, which makes it easy for anyone to improve
and contribute to the code\new{.}\old{: https://github.com/prisae/empymod}
\new{It comes with an extensive texting suite which is run automatically on
each commit, which guarantees some robustness and helps to avoid regression
bugs. Installation guidelines, documentation, and examples of its usage can be
found on its website at} \url{https://empymod.github.io}.%
\old{The documentation is hosted on https://empymod.readthedocs.io}
\cite{TLE.17.Werthmuller} gives a tutorial-style introduction to
\old{EM}\new{CSEM} modeling using empymod.

@@

The code consists of 5 files; these are 3 core modules plus \emph{utils}, which
contains input checks and other utilities, and \emph{filters}, containing the
FHT filter coefficients. The three core routines are: (1) \emph{kernel}, where
the wavenumber-domain calculation is carried out; (2) \emph{transform}, where
the Hankel and Fourier transforms are computed; and (3) \emph{model}, which
contains the actual modeling routines for end users. The main modeling routine
is \emph{bipole}, which can calculate frequency- and time-domain responses for
arbitrary oriented, electric or magnetic bipole sources and receivers of finite
length \new{(see also Appendix~\ref{apd:code})}.

The wavenumber-domain calculation in \emph{kernel} follows
\cite{GEO.15.Hunziker}, and calculates as such the complete wavefield for a
layered VTI model. The code makes no assumptions about the model: Source and
receiver can be placed anywhere in the model, including first and last layer,
and you have to define if the first layer is air or not. Bipoles can cross
layer boundaries. Depths, frequencies, and source-receiver configuration have
to be defined, and each layer is characterized with its horizontal resistivity
$\rho_\textrm{h}$, its electrical anisotropy $\lambda$, where $\lambda =
\sqrt{\rho_\textrm{v}/\rho_\textrm{h}}$, its horizontal and vertical magnetic
permeabilities $\mu_\mathrm{h}$ and $\mu_\mathrm{v}$, and the horizontal and
vertical electric permittivities $\epsilon_\mathrm{h}$ and
$\epsilon_\mathrm{v}$. The main difference between EMmod and empymod is the
Hankel transform, and a few additional fundamental differences: EMmod uses 2nd
order Bessel functions of the first kind $J_2$. Published FHT filters
generally provide coefficients for 0th and 1st order Bessel functions $J_0,
J_1$ only. \old{Therefore}\new{Consequently}, the recurrence relation
%
\begin{equation}
  J_2(kr) = \frac{2}{kr}J_1(kr) - J_0(kr)
  \label{eq:j2}
\end{equation}
%
is used, where $k$ and $r$ are the wavenumber and space-domain parameters,
respectively. Other differences are that empymod includes Fourier transforms,
hence time-domain calculation, and can model arbitrary rotated, finite bipoles.

@@

Calculation time can be reduced by using the splined version of the QWE and the
splined and lagged versions of the FHT. In addition, all time-consuming
calculations are set up to be able to run in parallel, with the help of the
Python-module \old{numexpr}\new{\texttt{numexpr}}. This can significantly speed
up calculations if you run big models with many layers and for many offsets and
frequencies. However, if you include empymod in an inversion scheme then it
might be better to parallelize the calls to empymod, instead of empymod itself.

It is important to note that calculations in the wavenumber domain depend
\emph{only} on offset $r = \sqrt{x^2+y^2}$; the scaling factor, which depends
on the angle $\varphi = \rm{atan2}(y, x)$, is multiplied afterwards. In order
to calculate for instance a circle around the source, the kernel has to be
called only once \new{for offset $r$}, and subsequently scaled by the
angle-dependent factor for each source-receiver pair. This makes the splined
version very powerful, as it can be used for irregularly distributed data%
\old{ as long as all receiver depths are the same.}%
\new{: you can calculate any amount of data points in the x-y-plane in one
single kernel call. This advantage fails as soon as the receivers have
different depths.}

@@

\new{Comparisons of empymod to other modelers using non-homogeneous models
follow in the ground-penetrating radar example and in the section
\emph{Comparison to Key (2009, 2012)}.}

@@

Table~\ref{tbl:analytical} shows run times for these models. A few notes worth
mentioning: empymod carries out a lot of input checks, as it is quite forgiving
in what you input. EMmod, on the other hand, reads the input file and writes
the result to an output file, and copies the first quadrant to the other three.
Both have \old{therefore}\new{then} some overhead, and the comparison is not
strictly 1:1. For the comparison the same model was used as shown above, on a
regular grid with a spacing of 100\,m. On the test-machine the standard QWE and
FHT empymod would run into memory issues for denser spacing, hence arrays of
more than some 10\,000s of offsets. The splined and lagged versions of EMmod
could handle it, however.
%
\tabl[btp]{analytical}{Run times for the half-space model shown in Figure~1, on
  a regular grid with spacing of 100\,m, for EMmod and different Hankel
  transform settings of empymod (times in milliseconds).}{
  \centering
  \begin{tabularx}{320pt}{rAAABBB}
  \toprule%
  EMmod & \multicolumn{3}{c}{QWE} & \multicolumn{3}{c}{FHT} \\
        & 1 & 2 & 3               & 1 & 2 & 3 \\
  \midrule%
   5040 & 10300 & 2530 & 1210 & 1480 & 875 & 6 \\
  \bottomrule%
  \end{tabularx}
}
%
QWE becomes faster for decreasing precision, not surprisingly, and the splined
and lagged version of FHT are faster than the standard version. The lagged FHT
is the fastest by quite a margin, and about 1/3 of that time is from the input
checks, so it takes only a few milliseconds to calculate the 11\,025 offsets.
What is interesting here is that the lagged FHT is still more precise than
EMmod with the given settings, which means a speed-up of a factor 1000 for the
same result.

Many 1D CSEM codes use the diffusive approximation that is valid for low
frequencies. The appealing part of the derivation of \cite{GEO.15.Hunziker} is
that it models the complete wavefield, hence it is also valid for high
frequencies and therefore wave-phenomena. As an extreme case,
\cite{GEO.15.Hunziker} show a 1D example for ground-penetrating radar with a
center frequency of 250\,MHz. To do so, they calculate 2048 frequencies in the
range of 1\,MHz to 2048\,MHz for the Fast Fourier transform from frequency to
time domain. Here I only calculated the FFT with 850 frequencies from 1\,MHz
to 850\,MHz and then zero-padded up to 2048\,MHz, for EMmod and empymod; tests
have shown that this is sufficient. Figure~\ref{fig:gpr} shows the model with
the survey setup and the four results for EMmod, \new{empymod with }QUAD for
relative and absolute tolerance of $10^{-12}$ and $10^{-20}$, respectively,
empymod with a 21\,pt QWE and relative and absolute tolerance of $10^{-10}$ and
$10^{-18}$, respectively, and empymod with FHT with the filter Key09-401,
together with the analytical solution for the arrival of the direct wave (red),
the wave refracted at the surface (yellow), and the wave reflected at the
subsurface interface (green).

EMmod and the QUAD and QWE transforms of empymod yield pretty much the same
result. It is only at later times that some noise can be seen, different in
each method. FHT shows the first arrivals well, however, afterwards the result
becomes extremely noisy.
%
\plot*[btp]{gpr}{width=1\textwidth}{GPR example for (b) EMmod, (c) QUAD
  for relative and absolute tolerance of $10^{-12}$ and $10^{-20}$,
  respectively, (d) empymod with 51\,pt QWE for relative and absolute tolerance
  of $10^{-8}$ and $10^{-15}$, respectively, and (e) empymod with FHT
  (Key09-401); run times are given in the subplot titles.}
%
Interesting are the calculation times for these three results. EMmod took
roughly 9.5 hours to calculate, empymod with QUAD about 7.8 hours, empymod with
QWE about 4.4 hours, and empymod with FHT under a minute. Note that QWE has
internally a check: If the kernel is too steep within given intervals, it uses
QUAD instead, as QWE cannot handle very steep functions. QWE uses in about 1/3
% (37%)
of the calculation of this example QUAD. Surprising is how good the FHT result
is, given that the filter was designed for CSEM data, hence frequencies 6 to 9
orders of magnitudes smaller and much bigger offsets. In any case,
EMmod/empymod are not optimized for nor intended to model GPR data, and
calculations in time-domain will be much faster and more precise. It is
nevertheless an interesting proof of concept and shows that EMmod/empymod model
the entire EM field. This leaves the question if it would be possible to
create a filter that works for the frequencies and the offsets required for GPR
calculations.

\new{Note that a common-midpoint (CMP) GPR-layout was chosen with increasing
offsets between the antennas, rather than the more common profiling layout with
fixed antenna separation, often with shielded antennas. The reason is that a
fixed offset calculation with shielded antennas of a 1D model would result in
exactly the same response at each measurement point, resulting in two flat
horizons, the direct and the reflected field. A CMP example is for this reason
more interesting than a profiling example.}

@@

\cite{GEO.12.Key} not only introduces QWE to geophysics, but also compares QWE
to FHT for various filters, some of which were specifically created for the
comparison. In order to calculate the models he translates parts of Dipole1D
\citep{GEO.09.Key} from Fortran to Matlab. The models are therefore isotropic
and use the diffusive approximation. He summarizes succinct and clear the FHT
method, outlines the QWE method, and made the codes available from the SEG
website.

@@

Table~\ref{tbl:runtimes} lists the run times for exactly the same models as
\cite{GEO.12.Key} compared in his Table~1: 1\,km water layer, followed by a
1\,km overburden of 1\,$\Omega\,$m, 100\,m reservoir of 100\,$\Omega\,$m, and
lastly 1 or 96 layers of underburden with a resistivity of 1\,$\Omega\,$m;
frequency is 1\,Hz, source depth 990\,m, and receivers are on the sea-bottom;
there are 1, 5, 21, 81, or 321 offsets between 0.5 and 20\,km. In this
comparison, a 9\,pt QWE is compared to a 201\,pt (Key12-201) and a 801\,pt
filter (Anderson82-801), where the absolute and relative tolerance are set so
that the QWE achieves a similar error-level as the filters ($\varepsilon_r =
10^{-6}$ for the standard QWE and $10^{-2}$ for the splined version,
$\varepsilon_a = 10^{-24}$ in both cases). In order to make a fair comparison I
re-run the script from \cite{GEO.12.Key}, and the run times on the test-machine
are roughly twice as fast as in the original paper. Next to it are the results
from empymod. It can be seen from the results that empymod is significantly
faster.%
%
\tabl[btp]{runtimes}%
{Run times comparing the codes from \cite{GEO.12.Key} (Key) and empymod (Wer)
for the same cases as in \cite{GEO.12.Key}, Table~1 (in milliseconds).}{
  \centering
\begin{tabularx}{1.05\textwidth}{ccYYWWYYWWYYWW}
  \toprule
  %
  & & \multicolumn{4}{c}{QWE: 9\,pt}&\multicolumn{4}{c}{FHT: 201\,pt filter}&
  \multicolumn{4}{c}{FHT: 801\,pt filter} \\
  %
  \cmidrule(rl){3-6} \cmidrule(rl){7-10} \cmidrule(rl){11-14}
  %
  & &
  \multicolumn{2}{>{\columncolor{MyGray}\arraybackslash}c}{Normal} &
  \multicolumn{2}{c}{Splined} &
  \multicolumn{2}{>{\columncolor{MyGray}\arraybackslash}c}{Normal} &
  \multicolumn{2}{c}{Lagged} &
  \multicolumn{2}{>{\columncolor{MyGray}\arraybackslash}c}{Normal} &
  \multicolumn{2}{c}{Lagged} \\
  %
  Lay& Off& Wer& Key& Wer& Key& Wer&  Key& Wer& Key& Wer&  Key& Wer& Key\\
  \midrule
  %
    5&   1&   7&  16&   2&  28&   1&   19&   1&  20&   2&   76&   2&  76\\
    5&   5&  13&  69&   4&  29&   3&   92&   1&  24&   7&  365&   3&  82\\
    5&  21&  19& 300&   6&  36&   8&  384&   2&  25&  25& 1522&   3&  83\\
    5&  81&  37&1154&  16&  62&  30& 1475&   2&  28& 100& 5865&   3&  86\\
    5& 321& 108&4581&  58& 165& 124& 5867&   2&  41& 437&23165&   3&  99\\
  \midrule
  100&  21& 118& 463&  13&  51&  92&  604&   8&  37& 304& 2402&  19& 127\\
  100&  81& 296&1792&  23&  75& 337& 2313&   8&  40&1234& 9215&  19& 129\\
  100& 321&1021&7226&  65& 178&1408& 9350&  10&  53&5448&36520&  19& 142\\
  %
  \bottomrule
\end{tabularx}}%
%
Important to note is, however, not the absolute speed of empymod, but the
difference between the various Hankel transform methods. The standard QWE
method, for a similar error level as FHT, is faster than the standard FHT mode
only for many offsets. More importantly, the lagged FHT is generally much
faster than any QWE. QWE is very useful to check the error level of a result.
If many kernel evaluations have to be carried out, and speed is of importance,
the lagged FHT is the preferred method. As \cite{GEO.12.Key} loops over each
wavenumber, the difference between the vectorized and non-vectorized version
can be best seen in the long 801\,pt filter.

@@

Table~\ref{tbl:runtimes2} lists the run times for the same model as before, but
for empymod and Dipole1D using the FHT method with the filter Key09-201, the
default filter in both Dipole1D and empymod. For the regular versions empymod
and Dipole1D perform very similar; empymod is a little faster for the smaller
tests, Dipole1D is slightly faster for the bigger tests. It can be seen from
the results that the parallel optimisation in empymod can result in a
significant speed-up. If the lagged convolution optimisation is chosen,
empymod is significantly faster than Dipole1D. In empymod, additional offsets
come at basically no cost in the lagged version, whereas Dipole1D still uses
more time to calculate more offsets. This is most likely due to the writing of
the output file, that becomes bigger with bigger offsets.
%
\tabl[btp]{runtimes2}
{Run times comparing Dipole1D \citep{GEO.09.Key} and empymod, using the filter
Key09-201 (times in milliseconds). Optimisation: [-] None, [par] parallel,
[spl] lagged FHT.}{ \centering
  \begin{tabularx}{.5\textwidth}{ccAAABB}
  \toprule
  %
     &    & \multicolumn{3}{c}{empymod} & \multicolumn{2}{c}{Dipole1D}\\
  %
  \cmidrule(rl){3-5} \cmidrule(rl){6-7}
  %
  Lay& Off&  - & par & spl &   - & spl\\
  \midrule
  %
    5&   1&    1&   2& 1&    4&  4\\
    5&   5&    3&   3& 2&    6&  5\\
    5&  21&    9&   5& 2&   12&  8\\
    5&  81&   30&  10& 2&   36& 18\\
    5& 321&  124&  34& 2&  130& 56\\
  \midrule
  100&  21&   91&  53& 9&   90& 13\\
  100&  81&  340& 106& 9&  333& 23\\
  100& 321& 1423& 313& 9& 1321& 62\\
  %
  \bottomrule
\end{tabularx}}%
%
Dipole1D was compiled using the free and open-source \emph{gfortran} compiler.
Compiling it with the (proprietary) \emph{ifort} compiler might result in
slightly faster run times. Note that Dipole1D always calculates all six
receiver components, whereas empymod only calculates the required component(s).
On the other hand empymod computes the whole, anisotropic wavefield, whereas
Dipole1D is isotropic and uses the diffusive approximation.

@@

\append[apd:code]{Code structure and basic usage}
